Static contract checking for Haskell

### 2011-08-19 paper.tex
Paper fix (Koen)
min version, first take. Shouldn't work
Paper update, mostly
Ttrans (datatype translation) becomes Ktrans (because we call the data constructors K anyway)
Strans (contract translation) becomes Ctrans (because s stood for satisfies but it's not obvious)

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ttrans}[1]{\mathcal{T} \llbracket #1 \rrbracket}
\newcommand{\strans}[1]{\mathcal{S} \llbracket #1 \rrbracket}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}

@@

\section{Translations}

@@

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
    \begin{array}{rccccccl}
      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
                                               &&&&&& \cdots & \\
                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
    \end{array}
    \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

\subsection{$\ttrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in \ref{ttrans}
$$\ttrans{\mbox{data } T = K_1, \dots, K_n} = S_1 \cup S_2 \cup S_3 \cup S_4$$

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(S_4)$ None of the $K_i$ is unreachable.
\item One may want to also state that if $\vec{x} \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\cf{\vec{x}} \to \cf{K_i(\vec{x})}$.
\end{itemize}

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

\subsection{$\strans{}$ -- Contracts}
We give in \ref{strans} the translation of contract satisfaction.
$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.

\begin{figure*}
\begin{eqnarray*}
  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
  \strans{e ::: \cfc} &=& \cf{e}
\end{eqnarray*}

\begin{eqnarray*}
  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
  \strans{e :/: \cfc} &=& ~\cf{e}
\end{eqnarray*}

$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$

\caption{$\strans{}$ -- Contract translation}
\label{strans}
\end{figure*}

@@

We say that the function implementation is correct wrt to its contract
iff $$\trans{M} \vdash \strans{f \in c}$$

@@

$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \strans{f_p \in c} \vdash \strans{f \in c}$$

@@

First, we want to check $f$'s contract. So we ask the theorem prover
if
$$ \dtrans{f}, \dtrans{g}, \strans{g} \vdash \strans{f} $$

@@

We should only include formulae that belongs to functions that are
actually used. For example, to prove $a$'s contract, we should only
include $f$ and $g$ translations, and we would ask equinox:
$$ \dtrans{f},\dtrans{g},\dtrans{a},\strans{f},\strans{g} \vdash
\strans{a}$$

\section{Correctness of the translation}
For the translation to be useful, $\trans{M} \vdash \strans{f \in c} $
should imply that $f \in c$.

>>>>

\newcommand{\etrans}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\dtrans}[1]{\mathcal{D} \llbracket #1 \rrbracket}
\newcommand{\ktrans}[1]{\mathcal{K} \llbracket #1 \rrbracket}
\newcommand{\ctrans}[1]{\mathcal{C} \llbracket #1 \rrbracket}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

@@

\section{Translations}
For an overview of the different translations we define, see \ref{trans}

\begin{figure}
  \label{trans}
  \begin{center}
  \begin{array}{rcl}
    \etrans{\mbox{expression}} &\to& t~(\mbox{Term}) \\
    \dtrans{f~\vec{x} = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \Phi~(\mbox{Set of formulae}) \\
    \ctrans{f \in c} &\to& \Phi~(\mbox{Set of formulae}) \\
  \end{array}
  \end{center}
  \caption{Translations}
\end{figure}

@@

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
    \begin{array}{rclcl}
      \forall \vec{a},y.~\min{y} \land f(\vec{a}) = y &\to& (\multicolumn{3}{c}{\min{e} \land (e = \bad \lor \bigvee_i \forall \vec{x}.~e = K_i(\vec{x}) \lor y = \unr)} \\      
      &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& y = \etrans{e_1} \\
      &\land& & \cdots & \\
      &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& y = \etrans{e_n}) \\
    \end{array}
    \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

\subsection{$\ktrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in \ref{ktrans}
$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
  on its $k$-th component.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ None of the $K_i$ is unreachable.
\item One may want to also state that if $\vec{x} \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\cf{\vec{x}} \to \cf{K_i(\vec{x})}$.
\end{itemize}

\begin{figure*}
  \label{ktrans}
  $$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
  where
  \begin{center}
    \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land K_i(\vec{x}) = a \land K_j(\vec{x}) = a) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~\min{a} \land a = K_i(\vec{x}) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array}
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
\end{figure*}

\subsection{$\ctrans{}$ -- Contracts}
We give in \ref{ctrans} the translation of contract satisfaction.
$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.

Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.

\begin{figure*}
\begin{eqnarray*}
  \ctrans{e \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{e} = \unr)\\
  \ctrans{e \in x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{e(x)}} \to (\ctrans{x \not \in c_1} \lor  \ctrans{\etrans{e(x)} \in c_2(x)})\\
  \ctrans{e \in \cfc} &=& \cf{\etrans{e}}
\end{eqnarray*}

\begin{eqnarray*}
  \ctrans{e \not \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = false \lor \etrans{e} = \bad)\\
  \ctrans{e \not \in x:c_1 \to c_2(x)} &=& \exists x.~ \ctrans{x \in c_1} \land  \ctrans{\etrans{e(x)} \not \in c_2(x)}\\
  \ctrans{e \not \in \cfc} &=& \lnot \cf{\etrans{e}}
\end{eqnarray*}

\begin{center}
\begin{array}{rcl}
 \ctrans{e \in c_1 \&\& c_2} &=& \ctrans{e \in c_1} \land \ctrans{e \in c_2} \\
 \ctrans{e \not \in c_1 \&\& c_2} &=& \ctrans{e \not \in c_1} \lor \ctrans{e \not \in c_2}\\
 \ctrans{e \in c_1 || c_2} &=& \ctrans{e \in c_1} \lor \ctrans{e \in c_2} \\
  \ctrans{e \not \in c_1 || c_2} &=& \ctrans{e \not \in c_1} \land \ctrans{e \not \in c_2}\\
 \ctrans{(a,b) \in (c_1,c_2)}   &=& \ctrans{a \in c_1} \land \ctrans{b \in c_2} \\
 \ctrans{(a,b) \not \in (c_1,c_2)}   &=& \ctrans{a \not \in c_1} \lor \ctrans{b \not \in c_2} \\
\end{array}
\end{center}

\caption{$\ctrans{}$ -- Contract translation}
\label{ctrans}
\end{figure*}

@@

We say that the function implementation is correct wrt to its contract
iff $$\trans{M} \vdash \ctrans{f \in c}$$

@@

$$\trans{M - f}, \dtrans{f=e[f/f_p]}, \ctrans{f_p \in c} \vdash \ctrans{f \in c}$$

@@

First, we want to check $f$'s contract. So we ask the theorem prover
if
$$ \dtrans{f}, \dtrans{g}, \ctrans{g} \vdash \ctrans{f} $$

@@

We should only include formulae that belongs to functions that are
actually used. For example, to prove $a$'s contract, we should only
include $f$ and $g$ translations, and we would ask equinox:
$$ \dtrans{f},\dtrans{g},\dtrans{a},\ctrans{f},\ctrans{g} \vdash
\ctrans{a}$$

\section{Correctness of the translation}
For the translation to be useful, $\trans{M} \vdash \ctrans{f \in c} $
should imply that $f \in c$.
